Interisland movements and oceanic swimming of woodland caribou in Newfoundland

### 2020-04-02 paper/manuscript.Rmd
minor edits

---
title: Interisland movements and oceanic swimming of woodland caribou in Newfoundland
author: Quinn M.R. Webber^1\*^, Jack G. Hendrix^1^, Alec L. Robitaille^2^,  Eric Vander Wal^1,2^
date: 
    ^1^Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial
   University of Newfoundland
    ^2^Department of Biology, Memorial University of Newfoundland, Canada
    #\*Corresponding author: Quinn M.R. Webber (webber.quinn@gmail.com)
output: 
    github_document
  # bookdown::pdf_document2:
  #    number_sections: true
  #    toc: false
  # bookdown::word_document2:
  #  reference_docx: template.docx
bibliography: ../references/swim.bib
#csl: csl.csl
---

@@

Oceanic swimming is relatively common among a minority of marked caribou. Of the
three collared individuals that swam the most, two were collared on Western
Indian Island and the third was collared on Fogo Island. Inter-island distances
between Western and Eastern Indian Islands, and the smaller islands around them,
are shorter than those between Fogo Island and others (Figure 2). It is possible
that caribou display natal philopatry [e.g. @Larue2018] to some of the smaller
islands and thus are more likely to swim between them more frequently than
animals that were born, ospend most of their time, on Fogo Island. The forage
limitation hypothesis is generally thought to apply uniformly across
populations, but individual caribou could evaluate the costs and benefits of
swimming to another island differently. It is possible that some individuals, in
our case ~10%, considered swimming to be less costly or perceived competition or
predation more acutely than their conspecifics.

For several decades after their introduction to Fogo Island, the caribou
population did not exceed ~100 individuals [@Bergerud1989, Newfoundland and
Labrador Wildlife Division, unpublished data]. During the 1990s, population
density reached 300 animals and it is unknown when caribou began to occupy other
islands, although anecdotal evidence suggests it was during the 1990s. We surmise that as
competition among conspecifics increased along with population size,
density-dependent habitat selection resulted in expansion of the population to
nearby islands. Following Ideal Free Distribution theory, caribou should swim to
new islands when the average fitness of individuals on the starting island
exceeds the density-fitness equilibrium [@Morris1987]. While this is an ultimate
explanation, our data suggest caribou only remain on smaller islands for a few
days at a time, so the fitness equilibrium remains a theoretical construct. More
likely, however, is that fine-scale competition, density-dependent habitat
selection, and forage depletion drives individuals to peridiodically swim
between islands. Given the presumed low density of coyotes on Fogo Island and
the minimal risk they pose, predation risk does not appear to be a major driver
of swimming.

>>>>

---
title: Interisland movements and oceanic swimming of woodland caribou in Newfoundland
author: Quinn M.R. Webber^1\*^, Jack G. Hendrix^1^, Alec L. Robitaille^2^,  Eric Vander Wal^1,2^
date: 
    ^1^Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial
   University of Newfoundland
    ^2^Department of Biology, Memorial University of Newfoundland, Canada
    #\*Corresponding author: Quinn M.R. Webber (webber.quinn@gmail.com)
output: 
  #  github_document
  # bookdown::pdf_document2:
  #    number_sections: true
  #    toc: false
   bookdown::word_document2:
    reference_docx: template.docx
bibliography: ../references/swim.bib
csl: ecology.csl
---

@@

Oceanic swimming is relatively common among a minority of marked caribou. Of the
three collared individuals that swam the most, two were collared on Western
Indian Island and the third was collared on Fogo Island. Inter-island distances
between Western and Eastern Indian Islands, and the smaller islands around them,
are shorter than those between Fogo Island and others (Figure 2). It is possible
that caribou display natal philopatry [e.g. @Larue2018] to some of the smaller
islands and thus are more likely to swim between them more frequently than
animals that were born, or spend most of their time, on Fogo Island. The forage
limitation hypothesis is generally thought to apply uniformly across
populations, but individual caribou could evaluate the costs and benefits of
swimming to another island differently. It is possible that some individuals, in
our case ~10%, considered swimming to be less costly or perceived competition or
predation more acutely than their conspecifics.

For several decades after their introduction to Fogo Island, the caribou
population did not exceed ~100 individuals [@Bergerud1989, Newfoundland and
Labrador Wildlife Division, unpublished data]. During the 1990s, population
density reached ~300 animals and it is unknown when caribou began to occupy other
islands, although anecdotal evidence suggests it was during the 1990s. We surmise that as
competition among conspecifics increased along with population size,
density-dependent habitat selection resulted in expansion of the population to
nearby islands. Following Ideal Free Distribution theory, caribou should swim to
new islands when the average fitness of individuals on the starting island
exceeds the density-fitness equilibrium [@Morris1987]. While this is an ultimate
explanation, our data suggest caribou only remain on smaller islands for a few
days at a time, so the fitness equilibrium remains a theoretical construct. More
likely, however, is that fine-scale competition, density-dependent habitat
selection, and forage depletion drives individuals to peridiodically swim
between islands. Given the presumed low density of coyotes on Fogo Island and
the minimal risk they pose, predation risk does not appear to be a major driver
of swimming.
